```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

set.seed(3620)  # reproducibility

```


```{r}
install.packages(c("tidyverse", "tidytext", "wordcloud", 
                   "RColorBrewer", "tm", "viridis", 
                   "GGally", "scales"))

library(tidyverse)
library(tidytext)
library(wordcloud)
library(RColorBrewer)
library(tm)
library(viridis)
library(GGally)
library(scales)
```

## Topic

Factors Affecting Airbnb Listing Prices
data source: https://insideairbnb.com
Inside Airbnb is a mission driven project that provides data about Airbnb's impact on residential communities

## Objectives

-   Cleaning and understanding Airbnb listing data\
-   Identify which factors most strongly influence price

## Skills Practiced

-   Dimension Reduction 
-   Statistical Modelling
-   Statistical Reasoning and Interpretation


### Data Cleaning Strategy
  To prepare the Airbnb dataset for analysis, I applied a systematic cleaning process using a checklist of common data issues. Each listing was evaluated and tagged with specific removal reasons so that the impact of filtering could be summarized transparently.
  
  The following issues were examined:
  Missing or invalid values: listings with missing, zero, or negative prices, or missing key predictors (accommodates, bedrooms, minimum_nights, room_type).
  
  #Incorrect data types: prices and percentage variables were converted from text to numeric format, and categorical variables were converted to factors.
  
  #Logical inconsistencies: listings with unrealistic combinations (e.g., 0 bedrooms but large accommodation capacity) were flagged.
  
  #Extreme outliers: the top 1% of prices were treated as outliers to prevent a small number of luxury listings from dominating statistical models.
  
  #Justification:
  These steps ensure that the final dataset contains valid, comparable listings and improves the reliability of regression analysis by reducing missing data problems, data-entry errors, and extreme leverage points.
  
  #Transparency:
  Rather than filtering immediately, each rule creates a “reason” flag. A summary table reports how many observations were removed for each reason, making the cleaning process reproducible and transparent.

```{r cleaning-audit}
library(dplyr)
library(readr)
library(stringr)

clean0 <- Listings %>%
  transmute(
    id = as.character(id),
    price_raw = price,
    price = parse_number(price),

    host_response_rate = parse_number(host_response_rate)/100,
    host_acceptance_rate = parse_number(host_acceptance_rate)/100,
    host_listings_count = as.numeric(host_listings_count),

    property_type = as.factor(property_type),
    room_type = as.factor(room_type),

    accommodates = as.numeric(accommodates),
    bathrooms = as.numeric(bathrooms),
    bedrooms = as.numeric(bedrooms),
    beds = as.numeric(beds),

    minimum_nights = as.numeric(minimum_nights),
    maximum_nights = as.numeric(maximum_nights),

    description = description
  )

# Define rules (you can edit thresholds)
clean1 <- clean0 %>%
  mutate(
    reason_price_missing_or_nonpos = is.na(price) | price <= 0,
    reason_min_nights_missing = is.na(minimum_nights),
    reason_room_type_missing = is.na(room_type),
    reason_bedrooms_missing = is.na(bedrooms),
    reason_accommodates_missing = is.na(accommodates),

    # Example "logical inconsistency" rule:
    reason_inconsistent_capacity = !is.na(bedrooms) & !is.na(accommodates) &
      bedrooms == 0 & accommodates >= 4,

    # Example extreme outlier rule (top 1%):
    price_p99 = quantile(price, 0.99, na.rm = TRUE),
    reason_price_outlier = !is.na(price) & price > price_p99
  )

# Summarize removal reasons
removal_summary <- clean1 %>%
  summarise(
    total_rows = n(),
    removed_price_missing_or_nonpos = sum(reason_price_missing_or_nonpos, na.rm=TRUE),
    removed_min_nights_missing = sum(reason_min_nights_missing, na.rm=TRUE),
    removed_room_type_missing = sum(reason_room_type_missing, na.rm=TRUE),
    removed_bedrooms_missing = sum(reason_bedrooms_missing, na.rm=TRUE),
    removed_accommodates_missing = sum(reason_accommodates_missing, na.rm=TRUE),
    removed_inconsistent_capacity = sum(reason_inconsistent_capacity, na.rm=TRUE),
    removed_price_outlier = sum(reason_price_outlier, na.rm=TRUE)
  )

removal_summary

```

## Research Question

Which listing characteristics have the greatest impact on Airbnb prices?

```{r load-data}
library(readr)
library(tidyverse)
library(dplyr)


Listings <- read_csv("C:/Users/pries/OneDrive/Documents/listings.csv",
                 show_col_types = FALSE)

glimpse(Listings)
dim(Listings)
head(Listings, 15)

```

```{r}
library(tidyverse)
library(readr)

df <- Listings

sub <- df |>
  transmute(
    price = parse_number(price),

    host_response_rate = parse_number(host_response_rate) / 100,
    host_acceptance_rate = parse_number(host_acceptance_rate) / 100,
    host_listings_count = as.numeric(host_listings_count),

    property_type = as.factor(property_type),
    room_type = as.factor(room_type),

    accommodates = as.numeric(accommodates),
    bathrooms = as.numeric(bathrooms),
    bedrooms = as.numeric(bedrooms),
    beds = as.numeric(beds),

    minimum_nights = as.numeric(minimum_nights),
    maximum_nights = as.numeric(maximum_nights)
  ) |>
  filter(!is.na(price), price > 0)

glimpse(sub)
head(sub)
```


```{r simple-regression-data}
library(dplyr)
library(readr)

simple_df <- sub |>
  dplyr::select(
    price,
    accommodates,
    bedrooms,
    room_type,
    minimum_nights
  ) |>
  dplyr::filter(
    !is.na(price),
    !is.na(accommodates),
    !is.na(bedrooms),
    !is.na(room_type),
    !is.na(minimum_nights)
  ) |>
  dplyr::mutate(
    log_price = log1p(price),
    room_type = as.factor(room_type)
  )

glimpse(simple_df)
head(simple_df)
```


## Visualizing Why Used Log of Price Instead of Price

```{r}
library(ggplot2)

# Original price
ggplot(simple_df, aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  ggtitle("Distribution of Price") +
  xlab("Price") +
  theme_minimal()

# Log-transformed price
ggplot(simple_df, aes(x = log_price)) +
  geom_histogram(fill = "lightgreen", color = "black", bins = 30) +
  ggtitle("Distribution of Log(Price)") +
  xlab("Log(Price)") +
  theme_minimal()
```

```{r}
library(ggplot2)
library(tidyr)

par(mfrow = c(1,2))  # Show two plots side by side

# QQ-plot for price
qqnorm(simple_df$price, main = "QQ-Plot: Price")
qqline(simple_df$price, col = "red")

# QQ-plot for log(price)
qqnorm(simple_df$log_price, main = "QQ-Plot: Log(Price)")
qqline(simple_df$log_price, col = "blue")


```


## Word Cloud
```{r}
# Convert to tibble
Listings <- as_tibble(Listings)

# Remove NA descriptions
airbnb_clean_text <- Listings %>%
  filter(!is.na(description))

# Unnest words
word_data <- airbnb_clean_text %>%
  unnest_tokens(word, description)

# Remove stop words
data("stop_words")

word_data_clean <- word_data %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_detect(word, "[a-z]"),nchar(word) > 2)  # remove numbers/punctuation

```

```{r}
word_counts <- word_data_clean %>%
  count(word, sort = TRUE)

```

```{r}
set.seed(3660)

wordcloud(words = word_counts$word,
          freq = word_counts$n,
          min.freq = 20,
          max.words = 100,
          random.order = FALSE,
          scale = c(2.5, 0.5),
          colors = brewer.pal(8, "Dark2"))
```

## Data Cleaning Strategy

The following issues were systematically checked:

1. Missing values
2. Invalid or zero prices
3. Extreme outliers
4. Incorrect data types
5. Duplicated entries
6. Logical inconsistencies (e.g., 0 bedrooms but accommodates 10)


## Fitting the linear regression


```{r simple-linear-model}
lm_simple <- lm(
  log_price ~ accommodates + bedrooms + room_type + minimum_nights,
  data = simple_df
)

summary(lm_simple)
```

```{r simple-linear-model}
lm_simple <- lm(
  log_price ~ bedrooms * room_type,
  data = simple_df
)

summary(lm_simple)
```
