---
title: "Factors on Airbnb Pricing in Montreal"
author: "Priestly"

output: github_document
editor_options: 
  markdown: 
    wrap: 72
---
### RESEARCH QUESTION:
How do listing characteristics such as capacity, room type, and minimum night requirements influence Airbnb pricing in Montreal, and does the effect of bedrooms vary across different room types?

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

set.seed(2026)

```

```{r load-data}
library(readr)
library(tidyverse)
library(dplyr)


Listings <- read_csv("listings.csv",
                 show_col_types = FALSE)

glimpse(Listings)
dim(Listings)
head(Listings, 15)

```



###Data Cleaning Strategy

To prepare the Airbnb dataset for analysis, I applied a systematic
rule-based cleaning process using a checklist of common data issues.
Each listing was evaluated and tagged with specific removal reasons so
that the impact of filtering could be summarized transparently.

Cleaning checks performed The following issues were examined:

-Missing or invalid values: listings with missing, zero, or negative
prices, or missing key predictors (accommodates, bedrooms,
minimum_nights, room_type).

-Incorrect data types: prices and percentage variables were converted
from text to numeric format, and categorical variables were converted to
factors.

-Logical inconsistencies: listings with unrealistic combinations (e.g.,
0 bedrooms but large accommodation capacity) were flagged.

-Extreme outliers: the top 1% of prices were treated as outliers to
prevent a small number of luxury listings from dominating statistical
models.

Justification: These steps ensure that the final dataset contains valid,
comparable listings and improves the reliability of regression analysis
by reducing missing data problems, data-entry errors, and extreme
leverage points.

Transparency: Rather than filtering immediately, each rule creates a
“reason” flag. A summary table reports how many observations were
removed for each reason, making the cleaning process reproducible and
transparent.

```{r cleaning}
library(dplyr)
library(readr)
library(stringr)

clean0 <- Listings %>%
  transmute(
    id = as.character(id),
    price_raw = price,
    price = parse_number(price),

    host_response_rate = parse_number(host_response_rate)/100,
    host_acceptance_rate = parse_number(host_acceptance_rate)/100,
    host_listings_count = as.numeric(host_listings_count),

    property_type = as.factor(property_type),
    room_type = as.factor(room_type),

    accommodates = as.numeric(accommodates),
    bathrooms = as.numeric(bathrooms),
    bedrooms = as.numeric(bedrooms),
    beds = as.numeric(beds),

    minimum_nights = as.numeric(minimum_nights),
    maximum_nights = as.numeric(maximum_nights),

    description = description
  )
```

```{r}
clean1 <- clean0 %>%
  mutate(
    reason_price_missing = is.na(price) | price <= 0,
    reason_min_nights_missing = is.na(minimum_nights),
    reason_room_type_missing = is.na(room_type),
    reason_bedrooms_missing = is.na(bedrooms),
    reason_accommodates_missing = is.na(accommodates),

    #logical inconsistency:
    reason_inconsistent_capacity = !is.na(bedrooms) & !is.na(accommodates) &
      bedrooms == 0 & accommodates >= 1,
      log_price = log1p(price),
  )
```

```{r}
# Summarize removal reasons
removal_summary <- clean1 %>%
  summarise(
    total_rows = n(),
    removed_price_missing = sum(reason_price_missing, na.rm=TRUE),
    removed_min_nights_missing = sum(reason_min_nights_missing, na.rm=TRUE),
    removed_room_type_missing = sum(reason_room_type_missing, na.rm=TRUE),
    removed_bedrooms_missing = sum(reason_bedrooms_missing, na.rm=TRUE),
    removed_accommodates_missing = sum(reason_accommodates_missing, na.rm=TRUE),
    removed_inconsistent_capacity = sum(reason_inconsistent_capacity, na.rm=TRUE)
  )

removal_summary

```

### Produce the final cleaned dataset + show % removed

```{r final-clean}
sub <- clean1 %>%
  filter(
    !reason_price_missing,
    !reason_min_nights_missing,
    !reason_room_type_missing,
    !reason_bedrooms_missing,
    !reason_accommodates_missing,
    !reason_inconsistent_capacity
  ) 

cat("Rows before:", nrow(clean1), "\n")
cat("Rows after :", nrow(sub), "\n")
cat("Percent kept:", round(100*nrow(sub)/nrow(clean1), 1), "%\n")

glimpse(sub)

```

## Exploratory Data Analysis:

## Numeric summaries

```{r eda-summaries}
library(scales)

sub %>%
  summarise(
    price_min = min(price, na.rm=TRUE),
    price_med = median(price, na.rm=TRUE),
    price_mean = mean(price, na.rm=TRUE),
    price_max = max(price, na.rm=TRUE)
  )

```

### Plot Of price

```{r}
library(ggplot2)

ggplot(sub, aes(price)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(labels = dollar_format()) +
  labs(
    title = "Distribution of Airbnb Prices (Cleaned Data)",
    x = "Price (CAD)",
    y = "Number of listings"
  ) +
  theme_minimal(base_size = 12)


```

## Visualizing Why Used Log of Price Instead of Price

```{r}
library(ggplot2)

# Original price
ggplot(sub, aes(x = price)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  ggtitle("Distribution of Price") +
  xlab("Price") +
  theme_minimal()

# Log-transformed price
ggplot(sub, aes(x = log_price)) +
  geom_histogram(fill = "lightgreen", color = "black", bins = 30) +
  ggtitle("Distribution of Log(Price)") +
  xlab("Log(Price)") +
  theme_minimal()
```



```{r text-high-low, fig.width=12, fig.height=6, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(tidytext)
library(wordcloud)
library(RColorBrewer)

# 1) Build grouped text data
text_df <- sub %>%
  filter(!is.na(description), !is.na(price)) %>%
  mutate(price_group = if_else(price >= median(price, na.rm = TRUE),
                               "High price", "Low price")) %>%
  select(id, price_group, description)

# Tokenize + remove stopwords + keep sensible tokens
words <- text_df %>%
  unnest_tokens(word, description) %>%
  anti_join(stop_words, by = "word") %>%
  filter(str_detect(word, "^[a-z]+$"), nchar(word) > 2)

word_counts <- words %>%
  count(price_group, word, sort = TRUE)

top_high <- word_counts %>%
  filter(price_group == "High price") %>%
  slice_max(n, n = 50)

top_low <- word_counts %>%
  filter(price_group == "Low price") %>%
  slice_max(n, n = 50)

# Wordcloud visualization (side-by-side comparison)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
wordcloud(words = top_low$word, freq = top_low$n,
          max.words = 50, random.order = FALSE,
          scale = c(2.2, 0.6),
          colors = brewer.pal(8, "Dark2"))
title("Low price listings")

wordcloud(words = top_high$word, freq = top_high$n,
          max.words = 50, random.order = FALSE,
          scale = c(2.2, 0.6),
          colors = brewer.pal(8, "Dark2"))
title("High price listings")

par(mfrow = c(1, 1))
overall_counts <- words %>%
  count(word, sort = TRUE)

# Overall wordcloud (all listings combined)
wordcloud(words = overall_counts$word, freq = overall_counts$n,
          min.freq = 20, max.words = 100,
          random.order = FALSE,
          scale = c(2.5, 0.5),
          colors = brewer.pal(8, "Dark2"))
title("Overall (all listings)")

```
 
 
### Within-Group Word Proportions (Normalize for Group Size) 
```{r word-proportions}
library(dplyr)
library(tidyr)
# Normalize word frequencies within each price group

# Count words by group
word_counts <- words %>%
  count(price_group, word)

# Total word counts per group
group_totals <- word_counts %>%
  group_by(price_group) %>%
  summarise(total = sum(n), .groups = "drop")

# Compute within-group proportions
word_props <- word_counts %>%
  left_join(group_totals, by = "price_group") %>%
  mutate(prop = n / total)

head(word_props,1000)
table(text_df$price_group)
table(words$price_group)
```


### Identifying Highly Shared (Generic) Words Across Price Groups
```{r}
common_words <- word_props %>%
  select(price_group, word, prop) %>%
  pivot_wider(
    names_from = price_group,
    values_from = prop,
    values_fill = 0
  ) %>%
  mutate(min_prop = pmin(`High price`, `Low price`)) %>%
  arrange(desc(min_prop))

common_words %>%
  select(word, `High price`, `Low price`, min_prop) %>%
  slice_max(min_prop, n = 75)
```

```{r}
summary(common_words$min_prop)

quantile(common_words$min_prop, probs = seq(0,1,0.1))

hist(common_words$min_prop, breaks = 50)
```


### Creating an Airbnb-Specific Stopword List
## We removed words with min_prop ≥ 0.002, which corresponds to extremely high shared-frequency terms well above the 90th percentile.

```{r airbnb-stopwords}
airbnb_stop <- common_words %>%
  filter(min_prop >= 0.002) %>%
  pull(word)

# Number of words removed
length(airbnb_stop)

# Inspect first few removed words
head(airbnb_stop, 80)
```

```{r}
airbnb_stop <- common_words %>%
  filter(min_prop >= 0.002) %>%
  arrange(desc(min_prop))

# Display removed words with their proportions in both groups
airbnb_stop %>%
  select(word, `High price`, `Low price`, min_prop)


#Example
word_counts %>%
  filter(word == "montreal")


```

```{r clean-final-words}
airbnb_stop_words <- airbnb_stop$word

words_clean <- words %>%
  filter(!word %in% airbnb_stop_words)


# Compare token counts before and after filtering
cat("Rows before:", nrow(words), "\n")
cat("Rows after :", nrow(words_clean), "\n")

```


### Constructing TF–IDF Text Features
```{r}
# Compute TF–IDF for each listing
dtm_tfidf <- words_clean %>%
  count(id, word) %>%
  bind_tf_idf(word, id, n)

# Inspect first few TF–IDF entries
head(dtm_tfidf,1000)
```

### Attach Price Labels to TF–IDF Features
```{r tfidf-attach-group}
dtm_tfidf_grouped <- dtm_tfidf %>%
  left_join(text_df %>% select(id, price_group), by = "id")

# Inspect the merged dataset
head(dtm_tfidf_grouped,40)
```


### Comparing Average TF–IDF by Price Group
```{r tfidf-compare}
avg_tfidf <- dtm_tfidf_grouped %>%
  group_by(price_group, word) %>%
  summarise(mean_tfidf = mean(tf_idf), .groups = "drop")

library(tidyr)

avg_compare <- avg_tfidf %>%
  pivot_wider(
    names_from = price_group,
    values_from = mean_tfidf,
    values_fill = 0
  ) %>%
  mutate(diff = `High price` - `Low price`) %>%
  filter(!is.na(diff))   # safety
head(avg_compare)

# Top words associated with High-price listings
top_high <- avg_compare %>%
  filter(diff > 0) %>%
  arrange(desc(diff)) %>%
  head(n = 75)

# Top words associated with Low-price listings
top_low <- avg_compare %>%
  filter(diff < 0) %>%
  arrange(diff) %>%
  head(n = 75)

# "Mid" words: closest to 0 difference (most similar usage)
mid <- avg_compare %>%
  mutate(abs_diff = abs(diff)) %>%
  arrange(abs_diff) %>%
  slice_head(n = 75) %>%
  select(-abs_diff)

head(avg_compare$diff)

# View only the words
top_high$word

# View first 20 words with their differences
top_high %>%
  select(word, diff) %>%
  head(20)


# View only the words
top_low$word

# View first 20 words with their differences
top_low %>%
  select(word, diff) %>%
  head(20)


# View only the words
mid$word

# View first 20 words with their differences
mid %>%
  select(word, diff) %>%
  head(20)

```



### Wordclouds of Price-Differentiating Language (TF–IDF Differences)
```{r tfidf-wordclouds, fig.width=18, fig.height=6}

library(wordcloud)
library(RColorBrewer)

par(mfrow = c(1,1),
    mar = c(2,2,3,2))

wordcloud(
  words = top_high$word,
freq = abs(top_high$diff),
min.freq = 0.5,  
  random.order = FALSE,
  scale = c(2.5, 0.7),
  colors = brewer.pal(8, "Dark2")
)

wordcloud(
  words = top_low$word,
  freq = abs(top_low$diff),
  min.freq = 0.5,  
  max.words = 150,
  random.order = FALSE,
  scale = c(2.5, 0.8),
  colors = brewer.pal(8, "Dark2")
)

wordcloud(
  words = mid$word,
  freq = (mid$diff),
  max.words = 75,
  random.order = FALSE,
  scale = c(2.5, 0.7),
  colors = brewer.pal(8, "Dark2")
)

par(mfrow = c(1,1))
```

## Dimension Reduction on Listing Descriptions

### Construct Document–Term Matrix (TF–IDF Weighted)
```{r dtm-matrix}

library(tidyr)
dtm_matrix <- dtm_tfidf %>% 
  select(id, word, tf_idf) %>% 
  pivot_wider( names_from = word, values_from = tf_idf, values_fill = 0 
               ) 

# Inspect matrix dimensions (listings × vocabulary size)
dim(dtm_matrix)
```


### Prepare Numeric Matrix for PCA
```{r}
dtm_numeric <- dtm_matrix %>%
  select(-id)

#confirm columns are numeric
is.numeric(dtm_numeric[[1]])
```


### Principal Component Analysis (Truncated SVD via irlba)
```{r}
# fast PCA for large matrices
if (!requireNamespace("irlba", quietly = TRUE)) install.packages("irlba")
library(irlba)

pca_model <- prcomp_irlba(as.matrix(dtm_numeric),
                          n = 10,      # number of PCs
                          center = TRUE,
                          scale. = TRUE)

# Examine proportion of variance explained
summary(pca_model)
 
```

### Variance Explained by Principal Components
```{r pca-variance}
imp <- summary(pca_model)$importance

# Display first 10 principal components
imp[, 1:10]
```


### Attach PCA Scores to Price Groups
```{r}
# 1) Build DTM (includes id column)
dtm_matrix <- dtm_tfidf %>% 
  select(id, word, tf_idf) %>% 
  tidyr::pivot_wider(names_from = word, values_from = tf_idf, values_fill = 0)

# 2) Make numeric matrix X with rownames = id
X <- dtm_matrix %>%
  mutate(id = as.character(id)) %>%
  tibble::column_to_rownames("id") %>%
  as.matrix()

# 3) Drop rows that are all zeros (no remaining words)
keep <- rowSums(X) != 0
X_use <- X[keep, , drop = FALSE]


library(irlba)
pca_model <- prcomp_irlba(X_use, n = 10, center = TRUE, scale. = TRUE)

# 5) Scores + correct ids
scores <- as.data.frame(pca_model$x)
scores$id <- rownames(X_use)

# 6) Attach price group
price_lookup <- text_df %>%
  distinct(id, price_group) %>%
  mutate(id = as.character(id))

scores <- scores %>%
  left_join(price_lookup, by = "id")

# sanity checks
nrow(scores)
table(scores$price_group, useNA = "ifany")
```


### Do PCA Language Components Differ by Price Group? (t-tests)
```{r pca-ttests}
# the ids used in PCA, in the same row order
ids_used <- dtm_matrix$id

scores <- as.data.frame(pca_model$x)
scores$id <- ids_used

# Attach price group labels
scores <- scores %>%
  mutate(id = as.character(id)) %>%
  left_join(price_lookup %>% mutate(id = as.character(id)), by = "id")
table(scores$price_group, useNA = "ifany")
head(scores$id)
head(price_lookup$id)

```


### Interpreting PC2 via Word Loadings
```{r}
loadings <- pca_model$rotation[,2]

# Top positive words (largest contributors to positive PC2 direction)
sort(loadings, decreasing = TRUE)[1:15]

# Top negative words (largest contributors to negative PC2 direction)
sort(loadings)[1:15]

```


### Create Structured Loading Table for PC2
```{r pc2-loadings-table}
# Extract PC2 loadings (word weights)
loadings_pc2 <- pca_model$rotation[, 2]

pc2_df <- tibble::tibble(
  word = rownames(pca_model$rotation),
  loading = as.numeric(loadings_pc2)
)

head(pc2_df)
```


### Identify Words Defining PC2 (Positive vs Negative Directions)
```{r}
# Top positive loadings (High price language)
top_high_pc2 <- pc2_df %>%
  arrange(desc(loading)) %>%
  slice_head(n = 75)

top_low_pc2 <- pc2_df %>%
  arrange(loading) %>%
  slice_head(n = 75)
  
# Inspect strongest contributors
head(top_high_pc2)
head(top_low_pc2)  
```


### Visualizing PC2 Language Structure
```{r pc2-wordclouds}
library(wordcloud)
library(RColorBrewer)

library(wordcloud)
library(RColorBrewer)

top_high_pc2 <- pc2_df %>% arrange(desc(loading)) %>% slice_head(n = 75)
top_low_pc2  <- pc2_df %>% arrange(loading)       %>% slice_head(n = 75)

top_high_pc2_plot <- top_high_pc2 %>% filter(abs(loading) > 0)
top_low_pc2_plot  <- top_low_pc2  %>% filter(abs(loading) > 0)

stopifnot(nrow(top_high_pc2_plot) > 0, nrow(top_low_pc2_plot) > 0)

par(mfrow = c(1,2), mar = c(1,1,3,1))
```

```{r}

```


------------------------------------------------------------------------

## 4) Statistical Modeling (principled selection + interactions + comparisons + assumptions)
We all know that before booking a reservation, there are certain key factors we look for. For example, we consider how many people a place can accommodate, the number of bedrooms and beds, the type of property, the price, and any minimum night requirements. These features naturally influence both the decision to book and the price of a listing.

Based on this, variables such as accommodates, bedrooms, beds, room type, and minimum nights were selected. We also prioritized variables with minimal missing data to ensure the reliability of the model.

```{r}
s1 <- lm(log1p(price) ~ accommodates, data=sub)
```
# Add bedrooms
```{r}
s2 <- lm(log1p(price) ~ accommodates + bedrooms, data=sub)
```
# Compare
```{r}
AIC(s1, s2)
```
Adding bedrooms reduces the AIC from 13810.18 to 13786.16, indicating an improvement in model fit.

# Add beds
```{r}
s3 <- lm(log1p(price) ~ accommodates + bedrooms + beds, data=sub)
AIC(s2, s3)
```
Including beds further reduces AIC to 13775.33, suggesting additional explanatory value.

# Add room_type
```{r}
s4 <- lm(log1p(price) ~ accommodates + bedrooms + beds + room_type, data=sub)
AIC(s3, s4)
```
Adding room type results in a substantial decrease in AIC (from 13775.33 to 12781.27), indicating that this variable has a very strong impact on price.

# Add minimum_nights
```{r}
s5 <- lm(log1p(price) ~ accommodates + bedrooms + beds + room_type + minimum_nights, data=sub)
AIC(s4, s5)
```
Including minimum nights further improves the model, reducing AIC to 12509.53.

Based on the forward selection process, the final model includes accommodates, bedrooms, beds, room type, and minimum nights, as it achieves the lowest AIC. This indicates that these variables collectively provide the best balance between model fit and complexity.

This forward selection approach provides a principled method for variable selection by systematically adding predictors and retaining them only when they improve model performance as measured by AIC.


### From Model selection

Pick variables based on: - (capacity, room type, bedrooms, minimum
nights) - plus **data quality** (few missing values)
```{r model-selection}
library(broom)

m1 <- lm(log1p(price) ~ accommodates + bedrooms + room_type + minimum_nights, data=sub)
m2 <- lm(log1p(price) ~ accommodates + bedrooms + beds + room_type + minimum_nights, data=sub)
m3 <- lm(log1p(price) ~ accommodates + bedrooms + room_type, data=sub)

AIC(m1, m2, m3) %>% arrange(AIC)



```

###Interactions A reasonable justification: - “The effect of bedrooms"
might differ by room type (entire home vs private room).”

```{r interaction-compare}
m_base <- m2
m_int  <- lm(log1p(price) ~ accommodates + beds + bedrooms*room_type + minimum_nights, data=sub)

AIC(m_base, m_int)     # model comparison


```

```{r baseline-model}
# Baseline additive model
m_base <- lm(
  log1p(price) ~ accommodates + bedrooms + beds +
    room_type + minimum_nights,
  data = sub
)

summary(m_base)
colSums(is.na(sub[, c("accommodates",
                      "bedrooms",
                      "beds",
                      "room_type",
                      "minimum_nights",
                      "price")]))


```

------------------------------------------------------------------------

# Interaction Multiple Linear Regression (Final Model)

> The effect of bedrooms on price may differ depending on room type
> (e.g., entire homes versus private rooms). To account for this
> possibility, an interaction between bedrooms and room type was
> included and compared with the baseline model.

```{r interaction-model}
# Interaction model
m_final <- lm(
  log1p(price) ~ accommodates + bedrooms + room_type + bedrooms*room_type +
    beds +
    minimum_nights,
  data = sub
)

summary(m_final)
```
### Handling Redundancy in the Interaction Model

We split the interaction to only include room types with variation in bedrooms, effectively excluding hotel rooms from the interaction. Comparing models using AIC shows identical values, indicating that this adjustment resolves redundancy without changing model fit.

```{r}
sub$bedrooms_private <- ifelse(sub$room_type == "Private room", sub$bedrooms, 0)
sub$bedrooms_shared  <- ifelse(sub$room_type == "Shared room", sub$bedrooms, 0)

m_fixed <- lm(log1p(price) ~ accommodates  + bedrooms + room_type + beds + minimum_nights+ 
                bedrooms_private + bedrooms_shared,
              data = sub)
summary(m_fixed)

AIC(m_final, m_fixed)
```
Overall, capacity and room type are the strongest drivers of price. Entire homes benefit more from additional bedrooms, while private rooms do not exhibit the same pricing pattern, highlighting important differences across listing types.